---
title: "Publishing runnable R chunks with CarmaR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Publishing runnable R chunks with CarmaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A 'Quarto' site rendered with the `carmar` filter keeps every page exactly as
rendered, and gives each visible R chunk a Run button that evaluates the
chunk on the reader's own computer, in the reader's own R. No server runs
the code; the page talks to a small local process - a CarmaR kernel - over a
loopback socket. This vignette is the author's story, then the reader's.

## The author

Enable publishing in the project once:

```{r}
carmar::use_publishing("path/to/project")   # or "." from inside the project
```

This copies the extension into `_extensions/carmar`, puts `- carmar` under
`filters:` in `_quarto.yml` (appended after any filters already there), and
writes the two options you may want to change when the file has none:

Prefer a folder to a command? Download `carmar-quarto.zip` from
<https://lacarm.com/carmar/> and unzip it inside the project: it creates the
same `_extensions/carmar/` folder. Then add `- carmar` under `filters:` in
`_quarto.yml` yourself (`quarto add https://lacarm.com/carmar/carmar-quarto.zip`
does the copy too).

```yaml
carmar:
  port: 4747
  label: Run on my computer
```

The port is the loopback port a reader's kernel will listen on; leave it at
4747 unless something else on readers' machines owns that port. The label is
the text of the Run button. A `_quarto.yml` that `use_publishing()` cannot
edit safely (an inline `filters: [...]` list) is left as it was and the lines
to add are printed.

Render and publish as you always do:

```sh
quarto render
quarto publish
```

Before publishing, or after, check the rendered output:

```{r}
carmar::check_publishing("_site")     # a book renders to "_book"
```

One row per page: `port` is `NA` for a page rendered without the filter, and
`assets` is `FALSE` when the page references the runtime script and
stylesheet but they are not beside it - which is what happens when a site is
uploaded without its `site_libs` directory.

A chunk that should be shown but never offered to run goes in a div with the
class `carmar-no-run`:

````markdown
::: {.carmar-no-run}
```{r}
unlink("~", recursive = TRUE)   # example only
```
:::
````

If the site sets its own Content Security Policy, allow the loopback socket
the page dials, on the port you chose:

```
connect-src ws://127.0.0.1:4747
```

## Pages that are already rendered

A report rendered years ago by R Markdown, a bookdown `_book/`, a pkgdown
article: none of them needs rendering again. One tag before `</head>` loads
the runtime, which gives every R block a Run button and the strip at the top:

```{r}
carmar::make_runnable("report.html")     # one file
carmar::make_runnable("_book")           # every .html under a folder, once
```

By default the page loads the runtime from CarmaR's published copy; pass
`assets = "local"` to copy the two files beside the pages instead. The tag is
inserted once; calling it again changes nothing.

## Any page at all

A reader can make a page runnable that its author never touched: a bookmark
whose address is the one-line script at <https://lacarm.com/carmar/> injects the
runtime on click. The kernel treats the site like any other — approved once per
session in the consent tab, code run only on Run — so no author cooperation is
needed, only a site whose Content-Security-Policy allows the script.

## The reader

The reader needs R and this package:

```{r}
install.packages("carmar")
carmar::listen("https://book.example")
```

`listen()` starts a headless kernel on the page's port and approves that one
site - scheme and host, nothing more - for the kernel's lifetime. It prints
one line:

```
CarmaR is listening on port 4747 for https://book.example - open the page and press Run.
```

Then the reader opens the page and presses Run. The chunk's output appears
under the chunk, from their own R, with their own packages and files.

If `listen()` is called without a site, the page's first Run opens a local
consent page naming the site, and the reader approves it there. Either way a
site is never trusted because it asked: approval is per exact origin, per
kernel, and ends when the kernel stops. The browser may also ask, once, for
permission for the site to reach the local network; that is the browser's
own question, and answering yes lets the page reach `127.0.0.1` and nothing
else.

This is execution, not a sandbox. What a reader runs has the same access as
their own R console, so authors keep the code visible and readers read a
chunk before running it.

To see and stop what is running:

```{r}
carmar::sessions()
carmar::stop_kernel(4747)
```
